Generation of TypeScript Declaration Files from JavaScript Code

### 2020-08-17 programming-2021/main.tex
Update results numbers and glob-to-regexp code example

declare function GlobToRegexp(glob: string, opts ?: GlobToRegexp.I__opts): RegExp;
declare namespace GlobToRegexp {
  export interface I__opts {
    'extended' ?: boolean;
    'globstar' ?: boolean;
    'flags'    ?: string;
  }
}
    \end{lstlisting}
  % \end{subfigure}
  \begin{center}
    \includegraphics[width=0.8\linewidth]{motivating-example-glob-to-regexp-vscode.png}
  \end{center}

@@

\begin{changethis}
  We implemented only basic TypeScript features and we focused our effort in building an
  end-to-end solution that generates valid and useful declaration files. Only the basic
  types \texttt{string}, \texttt{number}, \texttt{boolean} and its union were considered
  both for function parameters and interface properties. Common types such as
  \texttt{any}, arrays, function callbacks, tuples, intersections and features such as
  generics, index signatures were not implemented. Each declaration file was tagged with
  \texttt{dts-parse} and those containing unimplemented features were ignored. We add the
  \texttt{undefined} type for describing an optional parameter instead of adding the
  corresponding token \texttt{?}, which is also done by TypeScript in strict null checking
  mode\cite{typescript-optional-parameters-and-properties}.
\end{changethis}

@@

Figure \ref{fig:experiments-overall-funnel} shows that a declaration file was generated
for 244 modules out of 6029 modules and we identified 80 modules that have only the
features implemented by \texttt{dts-generate}. We obtained positive results for 69 of
them. Section~\ref{sec:experiments-evaluation} provides a detailed explanation of the overall
quality of the generated files.
Section~\ref{sec:experiments-declaration-files-generation} presents examples of the generated declaration files for
templates \textbf{module}, \textbf{module-class}, and \textbf{module-function}.

@@

\subsection{Evaluation}
\label{sec:experiments-evaluation}
We analyzed 82 of the 244 generated files. The remaining 162 files contained at least one
of the unimplemented TypeScript features so that they could not be considered. These
features were identified by \texttt{dts-parse} as tags.

As shown in \figref{fig:experiments-results-module-is-uuid}, the code examples have a
direct influence on the quality of the generated declaration file. We introduced the
concept of \texttt{solvable} and \texttt{unsolvable} differences in
\secref{sec:dts-compare}. Modules containing only \texttt{solvable} differences were
considered as positive results. We identified 72 such modules. We manually completed the
examples for 10 of those modules. Applying \texttt{dts-generate} with the completed
examples generates declaration files, which are equivalent to the files from DefinitelyTyped, as
shown in \figref{fig:experiments-results-manually-completed-examples}.
\begin{changethis}
  Expected differences are present: optional parameters are declared both with the actual
  and \texttt{undefined} type and the union type is handled through multiple overloads.
\end{changethis}
\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.80\linewidth}
    \begin{lstlisting}[language=TypeScript]
declare namespace gh {
  interface Options {
      enterprise?: boolean;
  }
  interface Result {
      user: string;
      repo: string;
      branch: string;
      https_url: string;
      tarball_url: string;
      clone_url: string;
      travis_url: string;
      api_url: string;
      zip_url: string;
  }
}

@@

We discovered only 10 modules with \texttt{unsolvable} differences. Most of them were
caused by the same error: 8 modules incorrectly declared an interface for the
\texttt{string} type. This interface contains standard properties and methods from the
JavaScript \texttt{String} object, such as \texttt{length} or \texttt{indexOf()}. The
remaining modules are \texttt{duplexer2} and \texttt{duplexer3}. Both of them rely on native
NodeJS types within the declaration file, which \texttt{dts-generate} could not identify
properly.

>>>>

declare function GlobToRegexp(glob: string, opts?: GlobToRegexp.I__opts): RegExp;
declare namespace GlobToRegexp {
  export interface I__opts {
    'extended'?: boolean;
    'globstar'?: boolean;
    'flags'?: string;
  }

}
    \end{lstlisting}
  % \end{subfigure}
  \begin{center}
    \includegraphics[width=0.8\linewidth]{motivating-example-glob-to-regexp-vscode.png}
  \end{center}

@@

\begin{changethis}
  We implemented only basic TypeScript features and we focused our effort in building an
  end-to-end solution that generates valid and useful declaration files. Only the basic
  types \texttt{string}, \texttt{number}, \texttt{boolean} and its union were considered
  both for function parameters and interface properties. Common types such as
  \texttt{any}, arrays, function callbacks, tuples, intersections and features such as
  generics, index signatures were not implemented. Each declaration file was tagged with
  \texttt{dts-parse} and those containing unimplemented features were ignored.
\end{changethis}

@@

Figure \ref{fig:experiments-overall-funnel} shows that a declaration file was generated
for 244 modules out of 6029 modules and we identified 60 modules that have only the
features implemented by \texttt{dts-generate}. We obtained positive results for 52 of
them. Section~\ref{sec:experiments-evaluation} provides a detailed explanation of the overall
quality of the generated files.
Section~\ref{sec:experiments-declaration-files-generation} presents examples of the generated declaration files for
templates \textbf{module}, \textbf{module-class}, and \textbf{module-function}.

@@

\subsection{Evaluation}
\label{sec:experiments-evaluation}
We analyzed 60 of the 244 generated files. The remaining 184 files contained at least one
of the unimplemented TypeScript features so that they could not be considered. These
features were identified by \texttt{dts-parse} as tags.

As shown in \figref{fig:experiments-results-module-is-uuid}, the code examples have a
direct influence on the quality of the generated declaration file. We introduced the
concept of \texttt{solvable} and \texttt{unsolvable} differences in
\secref{sec:dts-compare}. Modules containing only \texttt{solvable} differences were
considered as positive results. We identified 52 such modules. We manually completed the
examples for 10 of those modules. Applying \texttt{dts-generate} with the completed
examples generates declaration files, which are equivalent to the files from DefinitelyTyped, as
shown in \figref{fig:experiments-results-manually-completed-examples}.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.80\linewidth}
    \begin{lstlisting}[language=TypeScript]
declare namespace gh {
  interface Options {
      enterprise?: boolean;
  }
  interface Result {
      user: string;
      repo: string;
      branch: string;
      https_url: string;
      tarball_url: string;
      clone_url: string;
      travis_url: string;
      api_url: string;
      zip_url: string;
  }
}

@@

We discovered only 8 modules with \texttt{unsolvable} differences. Most of them were
caused by the same error: 7 modules incorrectly declared an interface for the
\texttt{string} type. This interface contains standard properties and methods from the
JavaScript \texttt{String} object, such as \texttt{length} or \texttt{indexOf()}. The remaining module is \texttt{stream}
remaining modules are \texttt{duplexer2} and \texttt{duplexer3}. Both of them rely on native
NodeJS types within the declaration file, which \texttt{dts-generate} could not identify
properly.
